A general and efficient representation of Ancestral Recombination Graphs

### 2023-08-25 paper.tex
Partial updates on precision of recomb info sectio

An important consequence of simplifying local trees to remove
unary nodes is that we lose information about recombination
events. This is related to the amount of \emph{precision} about
recombination events that we store, which is the topic of the next
section.

@@

% NEEDS more work!
\section*{Precision of ancestry information}
Ancestral recombination graphs are often characterised
as consisting of a complete record of the genetic history
of a sample. For example,
\cite{rasmussen2014genome} state that an ``ARG provides a record of all
coalescence and recombination events since the divergence of the sequences
under study'' and
according to \cite{deng2021distribution} an ARG
``provides all the information about the genealogical history of a sample,
including the locations of recombination events.'' It is worth
questioning, however, whether such comprehensive information
will always be available, and in particular whether our
representation of genealogical history should \emph{require}
such precision. Simulations will usually generate complete
information about the history of a sample, of course, but
we may not have sufficient information to infer such detail.

% The full ancestral recombination graph (ARG) is a structure that encodes all
% coalescence and recombination events resulting from the stochastic process of
% the coalescent with recombination.
\citet{brandt2021evaluation} define a ``full'' ARG as
``a structure that encodes all
coalescence and recombination events resulting from the stochastic process of
the coalescent with recombination'' (thus going against the general
historial trend discussed in the XXX section).
\citet{rasmussen2022espalier} have a less restrictive definition, in
which a ``full'' ARG is one that contains complete information about
recombination events, without necessarily being tied to a particular
stochastic process.
While [something positive], it is important to be clear on
what the overall goal of such precise inferences are,
what the limitations on when they can be made are,
and their downstream utility in applications.

% What is the goal of having fully precise ARGs?
% 1) studying the actual recombination events;
% 2) evaluating the likelihoods
The main reasons for wanting precise estimates of all recombination
events are to either compute a likelihood of the inferred ARG
under a statistical model, or to study the recombinants in
detail~\citep{rasmussen2022espalier}.
% Explain that we're really interested in the details of recombination
% itself with a small number of samples.
In the latter case, we might
have a small number of samples~\citep{guo2022recombination}.

Beyond cases in which visual inspection of the trees and the
effects of recombination is feasible, the main application
of having completely precise estimates of recombination events is
to evaluate the likelihood of the estimate under some
statistical model. The only statistical model available
is the coalescent with recombination
and its approximations~\citep{mcvean2005approximating,marjoram2006fast}.
In this case, the utility of a ``full`` ARG for some dataset
is limited by how well the dataset fits the assumptions
of the model.

% Not sure where this goes, but there are some critical points
% that we want to explain and not just throw out there in the
% discussion.
% A critical distinction between the eARG and gARG encodings
% is the level of precision about the time and location of
% recombination events that is required. In an eARG, we are
% required to be absolutely precise about every recombination
% [explain and contrast with gARG, referring to examples]

% JK: This is very rough, just jotting down the basic ideas,
% will need substantial revision.
It is not clear that such complete information will always
(or indeed ever) be available from inferences~\cite{shipilina2022origin}.
 We may be
able to precisely identify the location and timing
of recent recombinations (particularly if detailed pedigree
information is available) but beyond this, the information
required simply isn't present. When sampling from a
process such as the Sequentially Markov
Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
like ARGweaver~\citep{rasmussen2014genome}, it is true that
we do obtain a fully resolved ARG with precise
information about recombination, but it is reasonable
to question how much information from the data is used
to support older recombinations, in particular.

% This idea of a continuum of information about recombination
% being inferrable and representable is consistent with the ideas
% discussed by~\cite{shipilina2022origin}. There a haplotype block
% is synonymous with coalescent nodes in the ARG, which we can generalise
% as nodes in a gARG with at least two children. In principle, these
% nodes are identifiable by the mutations that occur on the outbound
% edges of the node.

It is helpful to consider the difference between fully
resolved binary trees and those that contain polytomies.

The situation is in fact directly analogous: an ARG that
contains fully resolved recombination information must have
at most parents per node.

At the very least, it is not reasonable for the \emph{data
structure} to require complete precision in order to
represent some less precise. This is a fundamental
advantage of the gARG encoding over the classical eARG
method: we are not forced to posit a precise cause (recombination
event) for every effect (difference between adjacent trees).
These differences in precision of recombination
information are illustrated in the ARGs estimated by different
methods, as we see in the next section.

In general, discussions tend to contrast an ARG
on one hand as holding \emph{complete} information about recombination
(and therefore between-tree correlation structure)
and a sequence of local trees on the other (with, by implication,
no direct information about between-tree correlations).
This is a false dichotomy:
a \emph{continuum} of information about recombination and between-tree
correlation structure can be represented and inferred, and all
such levels of information are useful, even if only for computational
efficiency.

To illustrate these points and to emphasise the heterogeneity
of the ARGs inferred by current methods, we inferred ARGs for the
the classical \citet{kreitman1983nucleotide} benchmark dataset
using four recent inference methods and show the results in
Fig~\ref{fig-inferred-args}.
While there is some consistency among the inferred ARGs (such as the samples
FR-F, WA-F and Af-F being grouped into the same clade), there
are also considerable differences. For this dataset, the minimum number
of recombination breakpoints required in the absence of recurrent mutation
is 7~\citep{song2003parsimonious}, and the number of inferred breakpoints
under the input mutation and recombination rates
varies from 2 (\relate) to 17 (\argweaver).
The ARGs inferred by the different methods also have major differences
in how recombination information is represented. Essentially,
\kwarg\ and \argweaver\ infer a specific causal recombination \emph{event}
for each breakpoint, and those events are fully
described in the output graph. On the other hand, \tsinfer\ and
\relate\ do not infer causal events but instead focus on
representing the inferred \emph{effect}: breakpoints between the
trees. It is important to note that this does not imply that
the methods must then necessarily infer only a sequence of
unrelated local trees
% More? See https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
% for discussion. I'm sure there were other examples of people saying
% that
(as claimed by e.g.~\citet{hejase2020summary}),
and information about recombination and the correlation structure
between trees is not ``all or nothing''.
Both methods capture correlation structure between trees:
\relate\ first infers local tree topologies and then identifies edges
that persist from one tree to the next as a separate step
(in a similar manner to \espalier~\citep{rasmussen2022espalier}), while
\texttt{tsinfer} explicitly constructs nodes that are shared across local
trees. See \cite{ignatieva2023distribution} for a detailed discussion
and comparsion of inference methods on this property.

% Because the ARGs inferred by \tsinfer\ and \relate\ do not contain
% explicit recombination events we cannot compute a likelihood
% under the coalescent with recombination (see appendix XXX).
% [Some stuff about whether we should see this as necessary or not,
% but maybe we should keep this for the conclusion?]

@@

\section*{Computational efficiency and data interchange}
ARG inference methods have been successfully applied to vast datasets,
including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
from UK Biobank~\citep{bycroft2018genome}
and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
Even larger datasets are
available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
and it seems inevitable that ARGs will be inferred for them.
The efficiency with which we can store and analyse these inferred
ARGs is therefore of great importantance, and a key factor
in the ultimate success of downstream ARG-based methods.
Another key factor in how important ARG-based methods become
across the wide range of predicted application
% TODO more?
areas~\citep{rasmussen2014genome,harris2019database,
hejase2020summary,schaefer2021ancestral,
fan2022genealogical,wohns2022unified,harris2023using}
is how well inference methods and downstream analysis applications
interoperate. Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems of converting data from one programs
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
While current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package
% FIXME-  more explicit?
(e.g.\ Relate, ARGneedle)
this approach scales poorly in terms of software development
effort, and is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse
software ecosystem.
% FIXME not sure where this goes
Earlier efforts to standardise ARG interchange
did not succeed~\citep{cardona2008extended,mcgill2013graphml}.

>>>>

An important consequence of simplifying ARGs to remove
unary nodes in local trees is that we lose some information
about recombination
events. This is related to the amount of \emph{precision} about
recombination events that we store, which is the topic of the next
section.

\section*{Precision of recombination information}
% The ARGs in Fig.~\ref{fig-simplification} represent different
% levels of detail about the same ancestral history. They represent
% the same set of recombination and common ancestor events,
As illustated in Fig.~\ref{fig-simplification}, successive levels
of ARG simplification reduce the amount of information about the
history of the sample that is stored. Some of the information lost,
for example, ``diamond'' removal (Fig.~\ref{fig-simplification}A),
seems like a reasonable tradeoff for a simpler structure.
The consequences of other simplifications, however, are
more subtle and relate directly to what can be known about
recombination events and the levels of precision that
we should seek to infer about them.

The ARGs in Fig.~\ref{fig-simplification} contain different
numbers of local trees ($6$, $5$, $5$ and $4$ for A through
D, respectively). When we move from A to B the local trees
for the intervals $(44,61]$ and $(61,87]$ are merged because
the only differences between them are their journeys through
nodes \noderef{l} and \noderef{m}. These nodes that participated
in the diamond are removed from the ARG, and we have lost
all information about the corresponding recombination at
position 61. Other nodes (e.g.\ \noderef{o} and \noderef{p})
have also been removed but these represent the \emph{parents}
of recombinants. The recombinant nodes themselves
(e.g. \noderef{n}) are still present, and represent precise
information about the time, genomic location and tree
branches involved
in the recombination event.

Fig.~\ref{fig-simplification}C has the same number of local trees
as Fig.~\ref{fig-simplification}B, but has less precise information
about recombination. Continuing the previous example, node
\noderef{n} has been removed from the graph because it was unary
in all of the local trees; its outbound edges to \noderef{s}
and \noderef{q} have effectively been ``pushed down''
to \noderef{e} (which is retained because it is the coalescent
parent of \noderef{a} and \noderef{b} over the interval
$(44, 100]$). We
have therefore lost precision about
the \emph{timing} of this recombination event, and know only
that it must have occured between the times of node \noderef{e}
and \noderef{q}.

Fig.~\ref{fig-simplification}D removes all unary nodes from the
local trees, and further reduces the precision of
recombination information. Node \noderef{e} has not been
removed from the graph because it is coalescent in the
final tree, but we no longer know that the recombination
event at position 30 was ancestral to it, or have
any indication of its timings. Furthermore,
trees for $(44, 87]$ and $(87, 100]$ were only distinguishable
by the passage of the former tree through nodes \noderef{e}
and \noderef{q}, and so the recombination on node \noderef{g}
at position 87 has been lost entirely.

% The example in Fig.~\ref{fig-simplification} is usually
% referred to as ``fully simplified'', and is the default output
% of simulation programs like \msprime. While there is significant
% precision about recombination events lost

The point of these examples is to emphasise that
information about recombination in an ARG is not ``all
or nothing'': there is a continuuum of detail
about recombination, in terms of genomic location, timing
the ancestral lineages involved, that we can describe.
While there has been a recent tendancy to emphasise the
importance of ``full ARGs''
\citep[e.g.][]{deng2021distribution,brandt2021evaluation,
rasmussen2022espalier}
which contain complete information about all recombination
events, it is reasonable to question the value of
such precise information and whether it is knowable.
[SOME MORE STUFF]

% Ancestral recombination graphs are often characterised
% as consisting of a complete record of the genetic history
% of a sample. For example,
% \cite{rasmussen2014genome} state that an ``ARG provides a record of all
% coalescence and recombination events since the divergence of the sequences
% under study'' and
% according to \cite{deng2021distribution} an ARG
% ``provides all the information about the genealogical history of a sample,
% including the locations of recombination events.'' It is worth
% questioning, however, whether such comprehensive information
% will always be available, and in particular whether our
% representation of genealogical history should \emph{require}
% such precision. Simulations will usually generate complete
% information about the history of a sample, of course, but
% we may not have sufficient information to infer such detail.

% % The full ancestral recombination graph (ARG) is a structure that encodes all
% % coalescence and recombination events resulting from the stochastic process of
% % the coalescent with recombination.
% \citet{brandt2021evaluation} define a ``full'' ARG as
% ``a structure that encodes all
% coalescence and recombination events resulting from the stochastic process of
% the coalescent with recombination'' (thus going against the general
% historial trend discussed in the XXX section).
% \citet{rasmussen2022espalier} have a less restrictive definition, in
% which a ``full'' ARG is one that contains complete information about
% recombination events, without necessarily being tied to a particular
% stochastic process.
% While [something positive], it is important to be clear on
% what the overall goal of such precise inferences are,
% what the limitations on when they can be made are,
% and their downstream utility in applications.

% % Not sure where this goes, but there are some critical points
% % that we want to explain and not just throw out there in the
% % discussion.
% % A critical distinction between the eARG and gARG encodings
% % is the level of precision about the time and location of
% % recombination events that is required. In an eARG, we are
% % required to be absolutely precise about every recombination
% % [explain and contrast with gARG, referring to examples]

% % JK: This is very rough, just jotting down the basic ideas,
% % will need substantial revision.
% It is not clear that such complete information will always
% (or indeed ever) be available from inferences~\cite{shipilina2022origin}.
%  We may be
% able to precisely identify the location and timing
% of recent recombinations (particularly if detailed pedigree
% information is available) but beyond this, the information
% required simply isn't present. When sampling from a
% process such as the Sequentially Markov
% Coalescent~\citep{mcvean2005approximating,marjoram2006fast}
% like ARGweaver~\citep{rasmussen2014genome}, it is true that
% we do obtain a fully resolved ARG with precise
% information about recombination, but it is reasonable
% to question how much information from the data is used
% to support older recombinations, in particular.

% % This idea of a continuum of information about recombination
% % being inferrable and representable is consistent with the ideas
% % discussed by~\cite{shipilina2022origin}. There a haplotype block
% % is synonymous with coalescent nodes in the ARG, which we can generalise
% % as nodes in a gARG with at least two children. In principle, these
% % nodes are identifiable by the mutations that occur on the outbound
% % edges of the node.

% It is helpful to consider the difference between fully
% resolved binary trees and those that contain polytomies.

% The situation is in fact directly analogous: an ARG that
% contains fully resolved recombination information must have
% at most parents per node.

% At the very least, it is not reasonable for the \emph{data
% structure} to require complete precision in order to
% represent some less precise. This is a fundamental
% advantage of the gARG encoding over the classical eARG
% method: we are not forced to posit a precise cause (recombination
% event) for every effect (difference between adjacent trees).
% These differences in precision of recombination
% information are illustrated in the ARGs estimated by different
% methods, as we see in the next section.

% In general, discussions tend to contrast an ARG
% on one hand as holding \emph{complete} information about recombination
% (and therefore between-tree correlation structure)
% and a sequence of local trees on the other (with, by implication,
% no direct information about between-tree correlations).
% This is a false dichotomy:
% a \emph{continuum} of information about recombination and between-tree
% correlation structure can be represented and inferred, and all
% such levels of information are useful, even if only for computational
% efficiency.

To illustrate these points and to emphasise the heterogeneity
of the ARGs inferred by current methods, we inferred ARGs for the
the classical \citet{kreitman1983nucleotide} benchmark dataset
using four recent inference methods and show the results in
Fig~\ref{fig-inferred-args}.
While there is some consistency among the inferred ARGs (such as the samples
FR-F, WA-F and Af-F being grouped into the same clade), there
are also considerable differences. For this dataset, the minimum number
of recombination breakpoints required in the absence of recurrent mutation
is 7~\citep{song2003parsimonious}, and the number of inferred breakpoints
under the input mutation and recombination rates
varies from 2 (\relate) to 17 (\argweaver).
The ARGs inferred by the different methods also have major differences
in how recombination information is represented. Essentially,
\kwarg\ and \argweaver\ infer a specific causal recombination \emph{event}
for each breakpoint, and those events are fully
described in the output graph. On the other hand, \tsinfer\ and
\relate\ do not infer causal events but instead focus on
representing the inferred \emph{effect}: breakpoints between the
trees. It is important to note that this does not imply that
the methods must then necessarily infer only a sequence of
\emph{unrelated} local trees (i.e., with either unlabelled internal
nodes or differently labelled nodes in each tree; see the
next section).
The ARGs inferred by \relate\ and \tsinfer\ both contain substantial
information about node sharing across trees, and hence capturing
the important correlation structures.

% \relate\ by first inferring local topologies and then identifies edges
% that persist from one tree, while
% local trees only emerge from \tsinfer\ as a consequence of the copying
% patterns inferred among ancestral haplotypes.

% % More? See https://github.com/tskit-dev/what-is-an-arg-paper/issues/38
% % for discussion. I'm sure there were other examples of people saying
% % that
% (as claimed by e.g.~\citet{hejase2020summary}),
% and information about recombination and the correlation structure
% between trees is not ``all or nothing''.
% Both methods capture correlation structure between trees:
% \relate\ first infers local tree topologies and then identifies edges
% that persist from one tree to the next as a separate step, while
% \texttt{tsinfer} explicitly constructs nodes that are shared across local
% trees.
% % Because the ARGs inferred by \tsinfer\ and \relate\ do not contain
% % explicit recombination events we cannot compute a likelihood
% % under the coalescent with recombination (see appendix XXX).
% % [Some stuff about whether we should see this as necessary or not,
% % but maybe we should keep this for the conclusion?]

@@

@@

\section*{Computational efficiency and data interchange}
ARG inference methods have been successfully applied to vast datasets,
including 500,000 genotyped humans~\citep{kelleher2019inferring,zhang2023biobank}
from UK Biobank~\citep{bycroft2018genome}
and over a million SARS-Cov-2 whole genomes~\citep{zhan2023towards}.
Even larger datasets are
available~\citep[e.g.][]{halldorsson2022sequences} % TODO more
and it seems inevitable that ARGs will be inferred for them.
The efficiency with which we can store and analyse these inferred
ARGs is therefore of great importantance, and a key factor
in the ultimate success of downstream ARG-based methods.
Another key factor in how important ARG-based methods become
across the wide range of predicted application
% TODO more?
areas~\citep{rasmussen2014genome,harris2019database,
hejase2020summary,schaefer2021ancestral,
fan2022genealogical,wohns2022unified,harris2023using,
nowbandegani2022extremely}
is how well inference methods and downstream analysis applications
interoperate. Without a well-defined and efficient shared format
for ARG data interchange,
users will suffer from problems of converting data from one programs
output to another, an unfortunate hallmark of population genetics
software for many years~\citep{excoffier2006computer}.
While current methods tend to tightly couple downstream analysis
of the inferred ARG with the inference
itself within the same software package
% FIXME-  more explicit?
(e.g.\ Relate, ARGneedle)
this approach scales poorly in terms of software development
effort, and is ultimately not compatible with the widespread use
of ARGs for routine data analysis, and a healthy and diverse
software ecosystem.
% FIXME not sure where this goes
Earlier efforts to standardise ARG interchange
did not succeed~\citep{cardona2008extended,mcgill2013graphml}.
